---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This paper explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{xcolor} 
  - \usepackage{hyperref}
  - \hypersetup{colorlinks = true, linkcolor = blue}

bibliography: references.bib 
editor_options: 
  markdown: 
    wrap: 72
---

\newpage 
\tableofcontents 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

The Gillespie Algorithm was presented in 1976 by Daniel T. Gillespie (@Gillespie1976). This algorithm was developed  to answer the question, “If a fixed volume V contains a spatially uniform mixture of N chemical species which can inter-react through M specified chemical reaction channels, then given the numbers of molecules of each species present at some initial time, what will these molecular population levels be at any later time?” (@Gillespie1977). Given the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).

Biochemical reactions in living cells are inherently stochastic. Reactions occur at random intervals, and the state of the system at any given time determines the probability of a particular chemical reaction occurring next. This stochasticity arises from the random thermal motion of molecules and the small numbers of reactant molecules involved. When molecule counts are low (often in the range of dozens or hundreds), random fluctuations in molecule collisions and interactions become significant, making reaction timing and outcomes probabilistic rather than deterministic. Consequently, deterministic approaches fail to accurately capture the dynamics of such systems, necessitating stochastic simulation methods like the Gillespie Algorithm.

In the following paper, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Topic Review section](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Theory section](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Application section](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006). These simulations will be performed using the R programming language (@citeR). Finally, our [Conclusion section](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.


# Topic Review {#sec-Topic-Review}

## Background

In this paper, we focus on the original version of the Gillespie Algorithm. This algorithm generates statistically correct trajectories to simulate chemical or biochemical reaction systems with high efficiency. By leveraging stochastic simulation, the Gillespie Algorithm enables researchers to model dynamics that cannot be solved analytically using the master equation, particularly for complex reaction networks. Its application is especially valuable in cases where deterministic methods fail to capture the inherent randomness of biochemical processes, such as systems with low molecular counts or highly interconnected reaction pathways.


# Theory {#sec-Theory}

## Master Equation

Generally we use the distribution function $P(S,t)$ to describe a many particle system at time t, where $P(S,t)$ is the probability for the system to be in the state $S$ at time $t$. 

The equation governing the evolution of $P(S,t)$ is called the master equation. During the time interval $[t,t+dt)$, $P(S,t)$ is changed due to two types of process:

1. The particles in the state $S$ leave due to some reaction $S\rightarrow S'$.

2. The particles in other state $S'$ enter the state $S$ state due to some reaction $S'\rightarrow S$.

Combining the two factors, we obtain the master equation with the form $^{[1]}$ $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$, where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.

The idea of master equation is inherently stochastic. Firstly $P(S,t)$ itself is a probabilistic description of the system. Next, in most of the cases, the reaction/process is stochastic. Therefore the master equation describes the ensemble average of all possible systems, here `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters. Following some certain initial distribution, we are not guaranteed to reach the same distribution at time $t$.

In the article `Cellular growth and division in the Gillespie algorithm` $^{[2]}$, the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$, where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.

With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$, here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.

## Gillespie Algorithm

The master equation is a coupled linear ordinary equation for $|S_Y|$ dependent variables $P(Y,t|Y_0,t_0)$, where $|S_Y|$ is the size of the state space. For a system with $n$ particles, $|S_Y|$ is of order $O(n^N)$. Therefore it is almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.

A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie $^{[3]}$. Using the idea that $P(Y,t|Y_0,t_0)$ describes the ensemble average of all possible micro-states given initial state $Y_0$ at time $t_0$, the Gillespie algorithm generates an ensemble of sample trajectories of the system with statistics which asymptotically converges to the solution of the corresponding Master Equation. 

In the Gillespie algorithm, we update the state of the system by determining 

(i) the time $\tau$ to the next reaction

(ii) which reaction $R_{\mu}$ will occur next. 

Assuming that each reaction $R_{\mu}$ is independent with rate $a_{\mu}=h_{\mu}c_{\mu}$, the first occurrence $\tau_{\mu}$ of the reaction $R_{\mu}$ follows the exponential distribution with rate/propensity $a_{\mu}$. The first occurrence time of some reaction is then $\tau=\min_{\mu}\tau_{\mu}$, it follows  the exponential distribution with an overall rate/propensity $A=\sum_{\mu}a_{\mu}$. Given that a reaction occurs, the probability that it is the $\mu'$-th reaction is $P(\mu=\mu')=\frac{a_{\mu'}}{A}$. Therefore the the original Gillespie recipe follows the following steps:

1) Input values $c_{\mu}$, $\mu=1,...,M$ and initial state $Y_0=(x_1,...,x_N)$ at $t_0$.

2) Compute the current propensities $a_{\mu}=h_{\mu}(Y)c_{\mu}$, $\mu=1,...,M$ and $A=\sum_{\mu}a_{\mu}$

3) Generate uniform random numbers $u_1,u_2\in[0,1)$ 

4) Compute the time interval $\tau$ until the next reaction according to distribution $\exp(A)$, i.e., $\tau=-\ln u_1/A$

5) Find the channel of the next reaction $\mu$, i.e., take $\mu$ to be the integer for which $\sum_{\nu=1}^{\mu-1}a_{\nu}<u_2*A\le \sum_{\nu=1}^{\mu}a_{\nu}$

6) Update time $t\rightarrow t+\tau$, and adjust $Y$ in accordance with the particular reaction $R_{\mu}$, i.e., update $Y\rightarrow Y+a_{\mu}$, and proceed to step 2 until $t$ reaches presetting terminal time $T_f$ and obtain a final state $Y_f$ at $T_f$.


After generating sufficiently many $N_{Traj}$ such trajectories, we can estimate $P(Y,T_f|Y_0,t_0)=\frac{\# Y_{f,Traj}=Y }{N_{Traj}}$.

The Gillespie algorithm is straightforward and easy to implement. However, it is only an approximate rather than an exact stochastic algorithm. A very large number of trajectories is needed to reach a reasonable accuracy. The complexity of the algorithm is very high and the accuracy is not guaranteed due to randomness. 

# Application {#sec-application}
## Photodimerization Simulation
We can simulate the dimerization of DNA pyrimidines, a molecular reaction which can lead to skin cancer. Bipyrimidine photodimerization occurs when UV radiation alters the chemical bonds of two consecutive pyrimidine bases, with two possible outcomes: the formation of the cyclobutene pyrimidine dimer (CPD) if a cycloaddition between the C5-C6 double bond of the pyrimidines occurs, or the formation of pyrimidine (6-4) pyrimidone (64-PP) if  a covalent bond forms between C4 and C6 of the pyrimidines (@MartinezFernandez2022).

Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 

We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per 10^6 normal bases) and 64-PP (~2 per 10^6 normal bases) found by exposing skin cells to a UVB dose of 0.2 J/cm^2 (@Mouret2006). As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs
c2 <- (37 / 1e6) * 10             # Formation of CPD
c3 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T2"],               # Reaction 1
    c2 * state["T1"] * state["T2"], # Reaction 2
    c3 * state["T2"]                # Reaction 3
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )
  
  # Record the initial state
  initial_row <- c(Time = time, state)
  output <- rbind(output, as.data.frame(t(initial_row), stringsAsFactors = FALSE))
  
  while (time < max_time) {
    # Calculate propensities
    props <- propensities(state)
    total_prop <- sum(props)
    
    # Stop if no reactions can occur
    if (total_prop == 0) break
    
    # Time to next reaction
    delta_t <- rexp(1, rate = total_prop)
    time <- time + delta_t
    
    # Select which reaction occurs
    reaction <- sample(length(props), 1, prob = props)
    
    # Update state based on reaction
    if (reaction == 1) state["T2"] <- state["T2"] - 1
    if (reaction == 2) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["CPD"] <- state["CPD"] + 1
    }
    if (reaction == 3) {
      state["T2"] <- state["T2"] - 1
      state["64PP"] <- state["64PP"] + 1
    }
    
    # Record the new state
    new_row <- c(Time = time, state)
    output <- rbind(output, as.data.frame(t(new_row), stringsAsFactors = FALSE))
  }
  
  # Convert columns to numeric for consistency
  output[] <- lapply(output, as.numeric)
  return(output)
}
# Run the simulation
max_time <- 100
simulation <- simulate_gillespie(state, max_time)
```

```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "Decay-Dimerization Reaction Set Simulation",
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```

## Radioactive Decay Simulation

This simulation models the stochastic decay of a single species (\( R \)), which represents a radioactive substance such as radon, radium, or plutonium. The decay process is governed by the following reaction:

\[
R \xrightarrow{c} 0
\]

where \( c \) is the decay rate constant that defines the likelihood of decay per unit time.

The Gillespie algorithm is used to simulate this system, capturing the inherent randomness of radioactive decay events. Unlike deterministic approaches, the Gillespie algorithm tracks individual decay events over time, providing an accurate representation of the stochastic nature of the process. The goal of this simulation is to observe the time evolution of the remaining radioactive molecules (\( R \)) and visualize their decay over time.

### Observations

The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process. Figure \@ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time. Figure \@ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.

#### General Explanation of Exponential Decay
The radioactive decay of \( R \) follows the well-known exponential decay law:

\[
R(t) = R_0 e^{-ct}
\]

where \( R_0 \) is the initial quantity of the substance, \( c \) is the decay constant, and \( t \) is time. As \( R \) decreases, the propensity function \( a(R) = c \cdot R \) ensures that the decay rate becomes slower over time. This results in a high number of decay events at the beginning of the simulation, which gradually diminishes as fewer molecules remain.

While the decay trajectory conforms to deterministic exponential decay theory, the Gillespie algorithm introduces stochastic variations. These arise because the time between successive reactions is governed by random exponential waiting times, and the occurrence of each reaction is probabilistic. As the number of \( R \) molecules declines, stochastic fluctuations become more apparent due to the reduced scale of the system.


```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 10000)  # Initial radioactive substance
max_time <- 1000  # Extended maximum simulation time

# Define propensity function
propensities <- function(state) {
  c * state["R"]
}

# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  time <- 0
  output <- data.frame(Time = numeric(0), R = numeric(0))
  output <- rbind(output, data.frame(Time = time, R = state["R"]))
  
  while (time < max_time) {
    # Calculate propensity
    prop <- propensities(state)
    if (prop == 0) break  # Stop if no reactions can occur
    
    # Time to next reaction
    delta_t <- rexp(1, rate = prop)
    time <- time + delta_t
    
    # Update state (decay reaction)
    state["R"] <- state["R"] - 1
    
    # Record the state
    output <- rbind(output, data.frame(Time = time, R = state["R"]))
  }
  
  return(output)
}

# Run the simulation
simulation <- simulate_gillespie(state, max_time)

```

#### Analysis of Figure \@ref(fig-decay-linear)
Figure \@ref(fig-decay-linear) illustrates the exponential decay of \( R \) over time. The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum. As \( R \) decreases, the curve flattens, reflecting the slower decay rate. This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.

The gradual flattening of the curve also underscores the diminishing probability of decay events as time progresses. This effect is a direct result of the propensity function \( a(R) = c \cdot R \), which decreases linearly with \( R \). The linear-scale plot effectively illustrates the overall trend of radioactive decay over time, making it particularly suitable for understanding the macroscopic behavior of the process.

```{r decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}

# Load ggplot2 for visualization
library(ggplot2)

# Linear-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "blue") +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (R)"
  ) +
  theme_minimal()
```

\newpage
#### Analysis of Figure \@ref(fig-decay-log)
Figure \@ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective. On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line. The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.

The logarithmic representation also highlights deviations from perfect linearity, which become more noticeable as the molecule count approaches zero. These deviations are the result of stochastic fluctuations inherent in the Gillespie algorithm, particularly in the later stages when \( R \) is low, and individual decay events have a greater impact on the overall trend. This logarithmic plot is especially useful for analyzing the detailed kinetics of the decay process and for confirming the exponential nature of the decay.

```{r decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
# Logarithmic-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (log(R))"
  ) +
  theme_minimal()

```

### Biological and Physical Significance

#### Radioactive Half-Life
The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes. The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay. It can be calculated as:

\[
t_{1/2} = \frac{\ln(2)}{c}
\]

For the decay constant \( c = 0.01 \) used in this simulation, the half-life is approximately \( t_{1/2} \approx 69.3 \) time units. This simulation demonstrates how the decay rate slows down over time as the radioactive substance depletes, with the time between successive decay events increasing in proportion to the decreasing molecule count.

#### Implications and Real-World Applications
Radioactive decay is a fundamental phenomenon in nuclear physics and has wide-ranging applications across multiple scientific disciplines. For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment. Similarly, in environmental science, the decay of isotopes such as radon is monitored to assess contamination levels and potential health risks.

The stochastic approach modeled here provides critical insights into systems with low molecule counts, where randomness plays a significant role. For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery. Likewise, in environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.

In the broader context of molecular reaction networks, the Gillespie algorithm's ability to capture both macroscopic trends and stochastic variations makes it an invaluable tool for studying systems with inherent randomness. By accurately simulating individual reaction events, it bridges the gap between theoretical models and real-world observations, offering a deeper understanding of complex kinetic processes.

# Conclusion {#sec-conclusion}
The Gillespie Algorithm represents a significant advancement in the numerical simulation of stochastic molecular reaction networks. Its ability to generate statistically valid trajectories of reaction systems provides critical insights into processes governed by inherent randomness, especially in systems with low molecule counts. This study has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes. 

The simulation of thymine photodimerization highlighted the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs. This has direct implications for understanding the molecular origins of UV-induced mutagenesis and skin cancer. Similarly, the radioactive decay simulation demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.

While the Gillespie Algorithm excels at modeling stochastic dynamics, it is not without limitations. The computational cost of simulating systems with large molecule counts or numerous reaction channels can become prohibitive, and its reliance on random number generation introduces variability that requires a sufficiently large number of trajectories for robust statistical analysis. Nevertheless, its utility in studying molecular systems with inherent randomness cannot be overstated.

The implications of this work extend beyond the specific applications discussed. By providing a framework for understanding and simulating stochastic reaction systems, the Gillespie Algorithm continues to play a pivotal role in advancing our knowledge of molecular biology, chemical kinetics, and physical chemistry. Future developments in computational efficiency and hybrid methods may further expand its applicability, enabling the simulation of even more complex systems with greater precision.


# References

[1]. N. G. V. Kampen, Stochastic Processes in Physics and Chemistry. North-Holland, Amsterdam, 1992.

[2]. T. Lu, D. Volfson, L. Tsimring, J. Hasty, Cellular growth and division in the Gillespie algorithm, Syst Biol (Stevenage). 2004 Jun;1(1):121-8


[3]. D. T. Gillespie, “Exact stochastic simulation of coupled chemical reactions,” J. Phys. Chem., vol. 81, no. 25, pp. 2340–2361,
1977.